\jaxcosmo: An End-to-End Differentiable and GPU Accelerated Cosmology Library

### 2022-11-21 paper_OJA.tex
Introducing in one-sentence the pmap and multi-mesh devel.

In JAX automatic vectorization is achieved using the \textit{jax.vmap} operation:
\begin{lstlisting}[language=iPython]
# Our function f only applies to scalars
def f(x):
	return y = 5 * x + 2
# Applying automatic vectorization
batched_f = jax.vmap(f)
# batched_f now applies to 1D arrays
\end{lstlisting}
Again, we stress that in this example, \texttt{batched\_f} will not be implemented in terms of a for loop, but with operations over vectors. The function above is trivial, but the same approach can be used to parallelize any function, from Limber integrals, to an entire likelihood.

@@

\section{Design of the \jaxcosmo\ library}
\label{sec-jaxcosmo-design}
\FrL{I'll do another pass over this section when I have internet, I want to compare to what we did in CCL.}

In this section, we describe the cosmological modeling provided by \jaxcosmo, and its implementation in JAX. The general design follows that of the Core Cosmology Library (CCL) \citep{2019ApJS..242....2C}, though \textbf{in its initial release, }
 \jaxcosmo\ only implements a subset of CCL features and options.

The cosmological parameters are organised in a JAX \textit{pyTree} container\footnote{\url{https://jax.readthedocs.io/en/latest/jax-101/05.1-pytrees.html}} \texttt{jc.Cosmology} where \texttt{jc} hereafter stands for \jaxcosmo. This container allows one to use \textit{autodiff} and \texttt{vmap} transparently \FrL{I don't think people know what autodiff and vmap are.}. The present set of exposed parameters is: $\Omega_{cdm}$ the cold dark matter density fraction, $\Omega_b$ the Baryonic matter density fraction, $h$ the Hubble constant divided by $100$~km/s/Mpc, $n_s$ the primordial scalar perturbation spectral index, $\sigma_8$ the variance of matter density perturbations at an $8$~Mpc/h scale, $\Omega_k$ the curvature density fraction, $w_0$ the first order term of dark energy equation and $w_a$ the second order term of dark energy equation of state. Notably, \texttt{jc.Planck15} instantiates a set of cosmological parameters inspired from the Planck 2015 results \citep{2016A&A...594A..13P}. 
\FrL{I'm not sure we want to keep this paragraph, it's not extremely useful. The only useful thing to discuss I guess might be the choice of parametrisation in terms of A or sigma8.}

@@

The computation of the evolution of the cosmological background follows a typical implementation of a Friedmann equation (\citealp[see e.g.][]{2005A&A...443..819P}):
\begin{equation}
    E^2(a) = \frac{\mathrm{H}^2(a)}{\mathrm{H}^2_0} = \Omega_m a^{-3} + \Omega_k a^{-2} + \Omega_{de} e^{f(a)}
\end{equation}
with $a=1/(1+z)$ the scale factor related to the redshift $z$, $H(a)=\dot{a}/a$ the Hubble parameter with $\mathrm{H}_0$ its present day value, $\Omega_m= \Omega_{cdm}+\Omega_b$, $\Omega_{de}=1-\Omega_k-\Omega_m$, and 
\begin{equation}
    f(a) = -3 (1 + w_0 + w_a) \ln(a) + 3 w_a (a - 1)
\end{equation}
Notably, the relativistic contributions of massless neutrinos and photon radiation, as well as the massive neutrino contribution are presently neglected. From these expressions, in \texttt{jc.background}, are computed the different cosmological distance functions. Notably, the radial comoving distance
\begin{equation}
     \chi(a) =  R_H \int_a^1 \frac{da^\prime}{{a^\prime}^2 E(a^\prime)}
     \label{eq:radial_comoving}
\end{equation}
with $R_H$ the Hubble radius. Using the scale factor ($a$) redshift ($z$) relationship, $\chi$ can be viewed as a function of $z$. 
%
\subsection{Growth of perturbations}
%
Currently, \jaxcosmo\  implements the \citet{Eisenstein_1998} transfer function and non-linear features through the \textit{halofit} model by \cite{2012ApJ...761..152T} or \cite{2003MNRAS.341.1311S} without the neutrino contribution introduced by \cite{10.1111/j.1365-2966.2011.20222.x}. No Baryon feedback modeling is considered yet. To perform \textit{autodiff} we have implemented various tools as differentiable ODEs using the Runge-Kutta algorithm, as well as Simpson quadrature and the Romberg algorithm for numerical integration. \FrL{Hummm it's a bit strange to talk about the theory and implementation in the same paragraph... Especially because we don't specify which concrete implementation is used here.}

@@

To make a bridge to these observables, two \textit{pyTree} objects are implemented: \texttt{NumberCounts} and \texttt{WeakLensing} tracers for galaxy clustering and weak lensing samples, with each instance of each representing a single tomographic redshift bin's number density \FrL{We probably should refer to those as Probes instead of Pytrees, in general we should separate the implementation details from the theory}. These tracers are implemented as two kernel functions:

@@

Notice that all described kernel functions can be also viewed as function of the scale factor $a$ or as function of the radial comoving distance $\chi$.

\FrL{We probably actually want to start by the Limber formula, which explicits what a kernel is, otherwise people don't necessarily know where these kernels come from.}
From these tracer objects, the angular power spectra $C_\ell^{ij}$ for the probes $(i,j)$ and for redshift bin window selections are computed in the first order Limber approximation \citep{PhysRevD.78.123506}:
%\JZ{Discuss the implementation of the Limber integral here. Maybe move down the bit in the growth of perturbations section?}
\begin{align}
    C_\ell^{i,j} \approx \left(\ell+\frac{1}{2}\right)^{m_i+m_j}\int\frac{d\chi}{c^2\chi^2}K_i(\chi)K_j(\chi)\,P\left(k=\frac{\ell+1/2}{\chi},z\right),\label{eq:Cell_limber}
\end{align}
with $P(k,z)$ the power spectra computed using the growth of perturbation models described above and a primordial power spectra scaling according to
\begin{equation}
    P(k) = A k^{n_s}
\end{equation}
The normalisation $A$ is computed using the $\sigma_8$ parameter as 
\begin{equation}
    A = \sigma_8^2 \times \left(\frac{1}{2 \pi^2} \int_0^\infty \frac{dk}{k} k^3 P(k) W^2(kR_8) \right)^{-1}
\end{equation}
\JEC{\st{At which z ?}}
with $R_8 = 8$ in unit of Mpc and $W(x)$ is related to the $j_1$ spherical Bessel function as
\begin{equation}
    W(x) = \frac{3j_1(x)}{x}
\end{equation}
Finally, the factors $m_i$ are equal to $(0,-2)$ for the \texttt{NumberCounts}, \texttt{WeakLensing} related kernels, respectively.

Remarkably, all the integrals involved in these computations can be  differentiated with respect to the cosmological parameters thanks to \textit{autodiff} and the implementation of the above mentioned integration quadrature \FrL{Are we using quadratures currently? I think by default we are using simple Simpson no?}\JEC{Correct: the current pip version uses Simpson but it is a quadrature even if it is qualified "simple".}.  An example is given in the context of DES Y1 3x2pts analysis (Sec~\ref{sec-DESY1}).
%
\subsection{Validation against the Core Cosmology Library (CCL)}
%
To illustrate the different features available with the present version of the library (\jaxcosmo\ \texttt{0.1rc9}, which is available in the Python Package Index PyPI\footnote{\url{https://pypi.org/}}), we have written a  companion notebook \nblink{CCL_comparison} to make a comparison with the well-established  Core Cosmology Library \citep{2019ApJS..242....2C}\footnote{\url{https://ccl.readthedocs.io}, version \texttt{2.5.1}.}. As examples, the Figures \ref{fig:chi_comparison},
\ref{fig:halofit_comparison} and \ref{fig:Cell_comparison}
show the radial comoving distance (Eq.~\ref{eq:radial_comoving}), the non linear matter power spectrum computation using the \texttt{halofit} fitting function and the angular power spectrum for galaxy-galaxy lensing (Eq.~\ref{eq:Cell_limber}) using the \texttt{NumberCounts} and \texttt{WeakLensing} kernel functions. \FrL{In addition, \jaxcosmo features a suite of validation tests against CCL, automatically validating the precision of all computations to within the desired numberical accuracy.} The relative differences between the two libraries are at the level of few $10^{-3}$ or better.

@@

\subsection{Instantaneous Fisher Forecasts}
\FrL{We should discuss why Fisher matrices are usually difficult first.}
We first illustrate, with a purely fictional case study, the computation of a Fisher matrix \citep{1997ApJ...480...22T,Stuart1991} using two methods with the \textit{autodiff} ability of JAX. For the detailed implementation, the reader is invited to look at the following companion notebook \nblink{Simple-Fisher}. In this example we use four tracer redshift distributions: two to define \texttt{WeakLensing} kernels and two for \texttt{NumberCounts} kernels. Then, the $10$ angular power spectra $C_\ell^{p,q}$ ($p,q:1,\dots,4$) are computed for $50$  logarithmically-spaced angular moments between $\ell=10$ and $\ell=1000$ using Equation \ref{eq:Cell_limber}. The Gaussian covariance matrix is computed simultaneously. A dataset is obtained from the computation of the $C_\ell^{p,q}$ with a fiducial cosmology. Then, the following snippet shows the log likelihood function $\mathcal{L}(\theta)$ implementation considering a constant covariance matrix ($\theta$ stands for the set of cosmological parameters). 
%\begin{minted}[fontsize=\footnotesize]{python}
\begin{lstlisting}[language=iPython]
@jax.jit
def likelihood(p):
    # Create a new cosmology at these parameters
    cosmo = jc.Planck15(Omega_c=p[0], sigma8=p[1])
    # Compute mean and covariance of angular Cls
    mu, C = jc.angular_cl.gaussian_cl_covariance_and_mean(cosmo, ell, tracers, sparse=True)
    # Return likelihood value assuming constant covariance, so we stop the gradient
    # at the level of the precision matrix, and we will not include the logdet term
    # in the likelihood
    P = jc.sparse.inv(jax.lax.stop_gradient(C))
    r = data - mu
    return -0.5 * r.T @ jc.sparse.sparse_dot_vec(P, r)
\end{lstlisting}
%\end{minted}
The \texttt{jc.sparse} functions are implementation of block matrix computations: a sparse matrix is represented as a 3D array of shape $(n_y, n_x, n_{diag})$ composed of $n_y \times n_x$ square blocks of size $n_{diag} \times n_{diag}$.  The \texttt{jax.jit} decorator builds a compiled version of the function at the first use.

@@

The NUTS sampler have been running with 16 chains of 1,000 samples each after a warm-up phase consisting of 200 samples, and the \texttt{max\_tree\_depth} has been set to 7 by default (ie. 128 steps for each iteration). \FrL{it should explain that vectorized means 16 chains in parallel on a single GPU.}
\JEC{The "vectorized" option runs multi chains in a single GPU thanks to the JAX \textit{vmap} mechanism. If one dispose of several GPUs then the "parallel" option using \textit{pmap}  mechanism runs one chain per GPU. As noted in \numpyro\ MCMC documentation, it is possible to mix parallel and vectorized sampling\footnote{Since version 0.9.0.}.}

The experiments have been undertaken on a single GPU either a NVidia Titan Xp (12GB RAM) on a desktop or an NVidia V100 (32GB RAM) at the IN2P3 Computing Centre\footnote{\url{https://cc.in2p3.fr/en/}}. The elapse time for these experiments was 20 hours.

>>>>

In JAX automatic vectorization is achieved using the \textit{jax.vmap} operation:
\begin{lstlisting}[language=iPython]
# Our function f only applies to scalars
def f(x):
	return y = 5 * x + 2
# Applying automatic vectorization
batched_f = jax.vmap(f)
# batched_f now applies to 1D arrays
\end{lstlisting}
Again, we stress that in this example, \texttt{batched\_f} will not be implemented in terms of a for loop, but with operations over vectors. The function above is trivial, but the same approach can be used to parallelize any function, from Limber integrals, to an entire likelihood. \JEC{Considering the case of multi devices (eg. GPU or TPU), JAX provides \texttt{pmap} which compile and execute in parallel replicates of the same code on each device, and recent experimental developments deal with parallelization of function over supercomputer-scale hardware meshes. In the examples detailed in this article, we have only relied on \texttt{vmap} functionality.}

@@

\section{Design of the \jaxcosmo\ library}
\label{sec-jaxcosmo-design}

In this section, we describe the cosmological modeling provided by \jaxcosmo, and its implementation in JAX. The general design follows that of the Core Cosmology Library (CCL) \citep{2019ApJS..242....2C}, though, in its initial release,
 \jaxcosmo\ only implements a subset of CCL features and options.

All \jaxcosmo\ data structures are organized as JAX \textit{pyTree} container objects, ensuring that two key JAX features are available to them: \textit{autodiff} and \textit{vmap}. The vmap feature enables any operation defined in JAX (including complicated composite operations) to be applied efficiently as a vector operation. The autodiff feature further makes it possible to take the derivative of any operation, automatically transforming a function that takes $n$ inputs and produces $m$ outputs into a new function that generates an $m \times n$ matrix of partial derivatives.

\jaxcosmo\ implements, for example, Runge-Kutta solvers for ODEs, and Simpson \& Romberg integration routines through this framework, so that we can automatically compute the derivative of their solutions with respect to their inputs. This includes the cosmological parameters (the standard set of $w_0 w_a CDM$ cosmological parameters is exposed, using $\sigma_8$ as an amplitude parameter), but also the other input quantities such as redshift distributions described below.

In the rest of this section we describe the cosmological calculations that are implemented using these facilities.

@@

The computation of the evolution of the cosmological background follows a typical implementation of a Friedmann equation (\citealp[see e.g.][]{2005A&A...443..819P}):
\begin{equation}
    E^2(a) = \frac{\mathrm{H}^2(a)}{\mathrm{H}^2_0} = \Omega_m a^{-3} + \Omega_k a^{-2} + \Omega_{de} e^{f(a)}
\end{equation}
with $a=1/(1+z)$ the scale factor related to the redshift $z$, $\mathrm{H}(a)=\dot{a}/a$ the Hubble parameter with $\mathrm{H}_0$ its present day value, $\Omega_m= \Omega_{cdm}+\Omega_b$, $\Omega_{de}=1-\Omega_k-\Omega_m$, and 
\begin{equation}
    f(a) = -3 (1 + w_0 + w_a) \ln(a) + 3 w_a (a - 1)
\end{equation}
Notably, the relativistic contributions of massless neutrinos and photon radiation, as well as the massive neutrino contribution are currently neglected. From these expressions, in \texttt{jc.background}, are computed the different cosmological distance functions, such as the radial comoving distance
\begin{equation}
     \chi(a) =  R_H \int_a^1 \frac{da^\prime}{{a^\prime}^2 E(a^\prime)}
     \label{eq:radial_comoving}
\end{equation}
with $R_H$ the Hubble radius. 
%Using the scale factor ($a$) redshift ($z$) relationship, $\chi$ can be viewed as a function of $z$. 
%
\subsection{Growth of perturbations}
%
Currently, \jaxcosmo\  implements the \citet{Eisenstein_1998} transfer function $T$ which transforms the primordial matter power spectrum to its late-time non-linear value:
\begin{equation}
    P(k, z) = P(k, z=\infty) \cdot T(k, z; \Omega_m, \Omega_b, ...),
\end{equation}
through the \textit{halofit} model by \cite{2012ApJ...761..152T} or \cite{2003MNRAS.341.1311S} without the neutrino contribution introduced by \cite{10.1111/j.1365-2966.2011.20222.x}. No Baryon feedback modeling is considered yet.

The primordial power spectrum is modelled with the standard form:
\begin{equation}
    P(k) = A k^{n_s}.
\end{equation}

The normalisation $A$ is parametrized via $\sigma_8$ at $z=0$ as 
\begin{equation}
    A = \sigma_8^2 \times \left(\frac{1}{2 \pi^2} \int_0^\infty \frac{dk}{k} k^3 P(k) W^2(kR_8) \right)^{-1}
\end{equation}
with $R_8 = 8 \mathrm{Mpc}/h$ and $W(x)$ related to the $j_1$ spherical Bessel function as
\begin{equation}
    W(x) = \frac{3j_1(x)}{x}
\end{equation}

@@

The angular power spectra $C_\ell^{ij}$ for the probes $(i,j)$ and for redshift bin window selections are computed in the first order Limber approximation \citep{PhysRevD.78.123506}:
\begin{align}
    C_\ell^{i,j} \approx \left(\ell+\frac{1}{2}\right)^{m_i+m_j}\int\frac{d\chi}{c^2\chi^2}K_i(\chi)K_j(\chi)\,P\left(k=\frac{\ell+1/2}{\chi},z\right),\label{eq:Cell_limber}
\end{align}
The factors $m_i$ are equal to $(0,-2)$ for the galaxy clustering and weak lensing respectively, and the functions $K(z)$ each represent a single tomographic redshift bin's number density. These tracers are implemented as two kernel functions:

@@

% Notice that all described kernel functions can be also viewed as function of the scale factor $a$ or as function of the radial comoving distance $\chi$.

% \FrL{We probably actually want to start by the Limber formula, which explicits what a kernel is, otherwise people don't necessarily know where these kernels come from.}

Because, like the other ingredients, the kernels are implemented as pyTree objects, \texttt{NumberCounts} and \texttt{WeakLensing}, all the integrals involved in these computations can be  differentiated with respect to the cosmological parameters  or the number densities using \textit{autodiff} and \jaxcosmo's implementation of integration quadrature. An example is given in the context of DES Y1 3x2pts analysis (Sec~\ref{sec-DESY1}).
%
\subsection{Validation against the Core Cosmology Library (CCL)}
%
To illustrate the different features available with the present version of the library (\jaxcosmo\ \texttt{0.1rc9}, which is available in the Python Package Index PyPI\footnote{\url{https://pypi.org/}}), we have written a  companion notebook \nblink{CCL_comparison} to compare it to the well-validated  Core Cosmology Library \citep{2019ApJS..242....2C}\footnote{\url{https://ccl.readthedocs.io}, version \texttt{2.5.1}.}. As examples, Figures \ref{fig:chi_comparison},
\ref{fig:halofit_comparison} and \ref{fig:Cell_comparison}
show the radial comoving distance (Eq.~\ref{eq:radial_comoving}), the non-linear matter power spectrum computation, and the angular power spectrum for galaxy-galaxy lensing (Eq.~\ref{eq:Cell_limber}) using the \texttt{NumberCounts} and \texttt{WeakLensing} kernel functions. \jaxcosmo\ features a suite of validation tests against CCL, automatically validating the precision of all computations to within the desired numberical accuracy; the relative differences between the two libraries are at the level of few $10^{-3}$ or better.

@@

\subsection{Instantaneous Fisher Forecasts}
Fisher matrices are a key tool in cosmology forecasting and experimental planning. By computing the Hessian matrix of a likelihood with respect to its parameters, we can find a Gaussian approximation to a posterior, which is usually sufficient for comparing the constraining power of different experimental configurations. As noted above, computing Fisher matrices is notoriously error-prone, since finite difference approximations to likelihoods must be carefully tuned for convergence, and observable calculations can be numerically unstable. Autodiff can help avoid this challenge.

We first illustrate, with an artificial case study, the computation of a Fisher matrix \citep{1997ApJ...480...22T,Stuart1991} using two methods with the \textit{autodiff} ability of JAX. For the detailed implementation, the reader is invited to look at the following companion notebook \nblink{Simple-Fisher}. In this example we use four tracer redshift distributions: two to define \texttt{WeakLensing} kernels and two for \texttt{NumberCounts} kernels. Then, the $10$ angular power spectra $C_\ell^{p,q}$ ($p,q:1,\dots,4$) are computed for $50$  logarithmically-spaced angular moments between $\ell=10$ and $\ell=1000$ using Equation \ref{eq:Cell_limber}. The Gaussian covariance matrix is computed simultaneously. A dataset is obtained from the computation of the $C_\ell^{p,q}$ with a fiducial cosmology. Then, the following snippet shows the log likelihood function $\mathcal{L}(\theta)$ implementation considering a constant covariance matrix ($\theta$ stands for the set of cosmological parameters). 
%\begin{minted}[fontsize=\footnotesize]{python}
\begin{lstlisting}[language=iPython]
@jax.jit
def likelihood(p):
    # Create a new cosmology at these parameters
    cosmo = jc.Planck15(Omega_c=p[0], sigma8=p[1])
    # Compute mean and covariance of angular Cls
    mu, C = jc.angular_cl.gaussian_cl_covariance_and_mean(cosmo, ell, tracers, sparse=True)
    # Return likelihood value assuming constant covariance, so we stop the gradient
    # at the level of the precision matrix, and we will not include the logdet term
    # in the likelihood
    P = jc.sparse.inv(jax.lax.stop_gradient(C))
    r = data - mu
    return -0.5 * r.T @ jc.sparse.sparse_dot_vec(P, r)
\end{lstlisting}
%\end{minted}
The \texttt{jc.sparse} functions are implementation of block matrix computations: a sparse matrix is represented as a 3D array of shape $(n_y, n_x, n_{diag})$ composed of $n_y \times n_x$ square blocks of size $n_{diag} \times n_{diag}$.  The \texttt{jax.jit} decorator builds a compiled version of the function at the first use.

@@

The NUTS sampler have been running with 16 chains of 1,000 samples each after a warm-up phase consisting of 200 samples, and the \texttt{max\_tree\_depth} has been set to 7 by default (ie. 128 steps for each iteration). \FrL{it should explain that vectorized means 16 chains in parallel on a single GPU.}
\JEC{The "vectorized" option runs multi chains in a single GPU thanks to the JAX \textit{vmap} mechanism. If one dispose of several GPU devices, then the using the JAX paralellization mechanism (\textit{pmap}), it is possible to launch on each GPU device the "vectorized" sampling as done in the previous snippet, and get back all the MCMC chains.}

Although, experiments have been undertaken on a single GPU either a NVidia Titan Xp (12GB RAM) on a desktop or an NVidia V100 (32GB RAM) at the IN2P3 Computing Centre\footnote{\url{https://cc.in2p3.fr/en/}}. The elapse time for these experiments was 20 hours.
